On the limits of making programming easy

### 2024-02-08 paper.tex
Joel proofread

\qquad In this paper, we investigate the limits of making programming easy. We use a dialectic method to circumscribe the design space within which easier programming systems may exist. In doing so, we bring together ideas on open-source software, self-sustainable systems, visual programming languages, but also the analysis of limits by Fred Brooks in his classic "No Silver Bullet" essay. We sketch a possible path towards easier programming of the future, but more importantly, we argue for the importance of proto-theories as a method for tackling the original motivating basic research question.

@@

\section{Introduction}
In the Cambridge Computer Laboratory, in the area known as The Street in front of the lecture theaters, on the cold winter day of 1st December 2023, we encounter two computer scientists who arrived for the Mycroftfest symposium. They are Tomas Petricek, a former PhD student of Alan, and Joel Jakubovic, who is, at the time of our meeting, a PhD student of Tomas. Joel recalls how he enjoyed Alan's \emph{Concepts in Programming Languages} module and the two start discussing their research.

@@

\says[\T]{I tried to fix this by designing a context-aware programming language, that would track the context requirements, or \emph{coeffects} in the type system~\cite{petricek-2013-coeffects}. With coeffects, the type would tell you not just what your functions take and return, but also that some part of your code can run only on some of the sites.}

@@

\says[\T]{Well, I ended up looking mostly at other use cases (dataflow, liveness) and only ever built a prototype implementation of a programming language with coeffect support.\footnote{The prototype can be found, in the form of an interactive essay, at \url{https://tomasp.net}.} But if someone implemented a proper programming language based on the theory, then I believe it would make programming client-server web applications easier...}

@@

\caption{The contrast between programming langauges and programming systems. We know how to study \emph{languages} (a), but not how to study \emph{systems} (b). }
\label{fig:langsys}
\vspace{-0.5em}
\end{figure}

@@

\says[\J]{A programming language is an abstract formal entity. A programming system is a running, interactive and stateful, collection of software. A system may implement a language (or multiple languages). It includes a runtime, a compiler or an interpreter, but it also consists of graphical interfaces used to edit code, debug code and so on.}

\says[\T]{I'm still not convinced. Of course, you need to write your code and invoke the compiler in some way. That's why you also have a terminal and Emacs to do programming.}

\says[\J]{You are thinking about the problem through the lens of the \emph{languages} paradigm. The point is that the \emph{systems} paradigm gives you a different perspective, different problems and different research methods. Gabriel wrote about this in his ``Structure of a programming language revolution'' \cite{rpg-2012-revolution}, where he points out that there was a paradigm shift through the 1990s from thinking about systems to thinking about languages. This encouraged more formal academic work on programming, but Gabriel also explains how some good ideas about programming became unthinkable because of this shift.}

@@

\says[\T]{Brooks \cite{brooks-1987-silver} analysed the different kinds of complexity that is involved in building software. The \emph{essential complexity} is inherent in the real-world. It comes from the fact that our software has to encode the many (perhaps ad-hoc) rules of the human world. The \emph{accidental compexity} comes from imperfection in our tools and languages.}

@@

\says[\J]{Like buildings, I think we often build software by gradual adaptation. We may not always think about it that way, but we should and we should embrace this view. Very often, you have software that does \emph{almost} what you want, but not quite.}

@@

\says[\J]{Many computer scientists like to see programming as a formal mathematical activity, but this is just a metaphor and it has its limits. De Millo, Lipton and Perlis
pointed this out in 1979 \cite{demillo-1979-proofs}! They believed that formal verification of software is bound to fail, because the complexity of the formal entities \emph{computer scientists} work with is different than the complexity of the formal entities \emph{mathematicians} work with. A mathematical proof is probably correct because mathematicians excitedly go over them with their colleagues on a whiteboard, but no such social process accompanies formal proofs of software.\footnote{For a detailed and a refined account of the debate that followed the publication of the paper, as well as the historical context, see the book written by MacKenzie~\cite{mackenzie-2001-proof}.}}

@@

\says[\J]{Yes, those tools give you rapid feedback. Perhaps a surprisingly good example here is Excel, which also, to some extent, keeps track of the chain of cause to effect in that you can view the source of the formula and see what other cells were used to compute the result. The editing is somewhat limited in that you can only edit the current formula. There is no easy way to follow the chain of causes. But it is the best example I can think of.}

@@

\says[\J]{Let's stop to think about notations for a bit. We can understand them better by charting the scope of change that can be achieved through a particular notation against the difficulty of using the notation. Figure~\ref{fig:sub1} shows this for a hypothetical accounting system and a spreadsheet system.}

@@

\says[\J]{Right. This is already better than regular software, because there are appropriate notations for changes of multiple levels of difficulties, but it is not perfect...}

@@

\says[\J]{I do not have a perfect answer to this question, but you can think of the basic substrate as a sort of ``user interface physics'' within which specific visual notations for particular problems could be defined.\footnote{Constraint-based systems for specifying user interfaces, such as Cassowary \cite{borning-1997-constraints}, can be seen as related to this idea, although they focus more on relating elements for the end-user than on developing a foundational underlying substrate.} If we had that, then I think we would make a step towards the ideal substrate for programming. And the substrate needs to be talking about actual running materialized system, rather than a static pre-scription of it that is a computer program in a static programming language today.}

@@

\says[\J]{Does that shed any light on why programming langauge researchers are so focused on formalization and proofs?}

@@

\says[\J]{I do not understand how you are supposed to make general and unifying claims if you do no first start by looking at the concrete and specific!}

\says[\T]{Well, there are some counter-examples to this. For most of its history, particle physics consisted of two sub-cultures that Peter Galison refers to as the ``image'' and ``logic'' traditions \cite{galison-1997-image}. While the latter focused on statistical analysis of observations, the former attempted to capture ``golden events'' that show the decay of a single particle. Such concrete trace was then hugely influential in shaping the theory of particle physics. And in the case of city planning, Jane Jacobs made a similar argument \cite{jacobs-1961-death} in favor of looking at specific cases, which she calls ``unaverage clues''. An example is a chain of bookshops that always remain open until late, except for one branch in Brooklyn. The clue tells you something significant about that part of the city.}

\says[\J]{You need something in between. In order to do any formalisation or a user study, you need conceptual clarity on what you're studying and what you're trying to find out. I don't think we have any such clarity on many interesting aspects of programming systems.}

@@

\says[\T]{Interesting, this kind of reasoning seems to be quite common in software engineering. I can think of another example from a book about the history of anti-ballistic missile systems \cite{slayton-2013-arguments}. One of the arguments for why such system cannot be reliably built was based on comparing the rate of change in the environment and the rate of change of the system. The idea was that the US cannot build a reliable defence system, because changing the software is more work than changing some of the characteristic of the Soviet ballistic missile. Again, this is a somewhat mathematical argument that provides conceptual clarity.}

\says[\J]{Right. I think we can refer to such models that provide conceptual clarity \emph{proto-theories}. I would say that we need more thinking like this! They can be later refined to more formal modles, but not prematurely. I would also add that mathematical formulations like those above are not the only kinds of proto-theories we should be creating.}

@@

\says[\J]{The list is quite long and I do not think it is complete, but it includes some of the
  characteristics of programming systems that we have talked about alread. For example,
  \emph{modes of interaction} captures whether you do everything in a single environment or
  whether ``using'' is distinct from ``programming''; \emph{notational structure} captures whether
  there is just a single adaptable notation or a range of different ones; \emph{self-sustainability}
  captures the degree to which the system can be modified from within itself and so on.}

@@

\says[\T]{Didn't you link the shift from thinking about programming langauges to thinking about programming systems to the idea of Kuhnian scientific revolution before?}

\says[\J]{Well, yes. If this really was a scientific revolution, then the shift to thinking about programming systems would put more emphasis on research methods that can talk about them in their full complexity. It would not invalidate programming language theory, but it would certainly make it less relevant. However, we are quite far from what Kuhn called ``normal science''. It is still early days for programming systems research!}

@@

\says[\J]{Right. The full realisation of the aims of the framework is a long-term research program. What we have already is the seed to start it off. We have a set of dimensions that come with a qualitative description each. Future work, as set out towards the end of my dissertation, involves making these descriptions more precise and possibly even qualtifiable.}

@@

\says[\J]{That is certainly one task for the experimentalists. But they can also explore the design space to find new designs. To get back to our original motivation of making programming easier, one of the chapters in my thesis aims to explore programming systems that have three properties: \emph{notational freedom}, which would give you the kind of adaptable substrate we talked about,
\emph{explicit structure}, which is a pre-condition for building such substrate and
\emph{self-sustainability}, which would make it possible to evolve and improve the system from
within itself, enabling a virtuous cycle of self-improvement.}

@@

\says[\T]{Well, Joel was practicing the elevator pitch for his PhD thesis, but I suppose we got a bit distracted and ended up talking about whether it is possible for programming to be easier, whether there are any fundamental limits of that and also what kind of research methodology can get us there!}

@@

\says[\J]{On a more general level, we agreed that we need more thinking about interactive, stateful programming systems and their characteristics. Programming can only become easier if we start thinking about how programming systems are used. In other words, think more about \emph{programming systems} than about \emph{programming languages}. Today, this is tricky, because we lack the right vocabulary for talking abotu programming systems, but the technical dimensions is a useful \emph{proto-theory} that serves as the starting point!}

@@

\says[\J]{Well, we do not have a concrete system yet, but we believe it will have a number of general properties. It needs to be built around a programming substrate that can grow with the user. It should make it easy to make small changes to the system, but also allow making more complex and harder to make changes. We also need to think about programming as gradual adaptation, to overcome the Brooks' ``doom prophecy'' of essential complexity. But I remain optimistic.}

@@

\says[\J]{I would love to see more research that looks at how programming languages are embedded in broader stateful interactive systems. This is something where we are lacking both theory and new interesting design ideas. I also think programming researchers should look at things that we do not really see as programming languages. What is it that makes systems like Excel, but also Flash, Hypercard or game engines work well as programming tools? We also need to be more fearless in looking at problems that are interesting, but informal. If we are to study informal properties that are not simply true or false, we will also need more focus on finding consensus among researchers working on programming tools, such as agreeing on the right proto-theories on which to further advance our work.\footnote{A recent seminar also focused on the problem of developing new ``theories of programming'' \cite{latoza-2023-theories}. However, we believe that a theory needs to start as an informal proto-theory and that strong scientific framing in terms of hypotheses may inhibit progress.}}

>>>>

\qquad In this paper, we investigate the limits of making programming easy. We use a dialectic method to circumscribe the design space within which easier programming systems may exist. In doing so, we not only bring together ideas on open-source software, self-sustainable systems, visual programming languages, but also the analysis of limits by Fred Brooks in his classic "No Silver Bullet" essay. We sketch a possible path towards easier programming of the future, but more importantly, we argue for the importance of proto-theories as a method for tackling the original motivating basic research question.

@@

\section{Introduction}
It is a cold winter morning on 1st December 2023 outside the Cambridge Computer Laboratory. In the area known as The Street in front of the lecture theaters, on the cold winter day of 1st December 2023, we encounter two computer scientists who arrived for the MycroftFest symposium. They are Tomas Petricek, a former PhD student of Alan, and Joel Jakubovic, who is Tomas' PhD student at the time. Joel recalls how he enjoyed Alan's \emph{Concepts in Programming Languages} module and the two start discussing their research.

@@

\says[\T]{I tried to fix this by designing a context-aware programming language that would track the context requirements, or \emph{coeffects}, in the type system~\cite{petricek-2013-coeffects}. With coeffects, the type would tell you not just what your functions take and return, but also that some part of your code can run only on some of the sites.}

@@

\says[\T]{Well, I ended up looking mostly at other use cases (dataflow, liveness) and only ever built a prototype implementation of a  language with coeffect support.\footnote{The prototype can be found, in the form of an interactive essay, at \url{https://tomasp.net}.} But if someone implemented a proper language based on the theory, then I believe it would make programming client-server web applications easier...}

@@

\caption{The contrast between programming languages and programming systems. We know how to study \emph{languages} (a), but not how to study \emph{systems} (b). }
\label{fig:langsys}
\vspace{-0.5em}
\end{figure}

@@

\says[\J]{A programming \emph{language} is an abstract formal entity. A programming \emph{system} is a running, interactive and stateful collection of software. A system may implement a language (or multiple languages). It includes a runtime and a compiler or an interpreter, but it also consists of graphical interfaces used to edit code, debug code and so on.}

\says[\T]{I'm still not convinced. Of course you need to write your code and invoke the compiler in some way. That's why you also have a terminal and Emacs to do programming.}

\says[\J]{You are thinking about the problem through the lens of the \emph{languages} paradigm. The point is that the \emph{systems} paradigm gives you a different perspective, different problems and different research methods. Gabriel wrote about this in his ``Structure of a Programming Language Revolution'' \cite{rpg-2012-revolution}, where he points out that there was a paradigm shift through the 1990s from thinking about systems to thinking about languages. This encouraged more formal academic work on programming, but Gabriel also explains how some good ideas about programming became unthinkable because of this shift.}

@@

\says[\T]{Brooks \cite{brooks-1987-silver} analysed the different kinds of complexity involved in building software. The \emph{essential complexity} is inherent in the real-world. It comes from the fact that our software has to encode the many (perhaps ad-hoc) rules of the human world. The \emph{accidental compexity} comes from imperfection in our tools and languages.}

@@

\says[\J]{Like buildings, I think we often build software by gradual adaptation. We may not always think about it that way, but we should embrace this view. Very often, you have software that does \emph{almost} what you want, but not quite.}

@@

\says[\J]{Many computer scientists like to see programming as a formal mathematical activity, but this is just a metaphor and it has its limits. De Millo, Lipton and Perlis
pointed this out in 1979 \cite{demillo-1979-proofs}! They believed that formal verification of software is bound to fail, because the complexity of the formal entities \emph{computer scientists} work with is different than the complexity of the formal entities \emph{mathematicians} work with. A mathematical proof is probably correct because mathematicians excitedly go over it on a whiteboard with their colleagues, but no such social process accompanies formal proofs of software.\footnote{For a detailed and a refined account of the debate that followed the publication of the paper, as well as the historical context, see the book written by MacKenzie~\cite{mackenzie-2001-proof}.}}

@@

\says[\J]{Yes, those tools give you rapid feedback. Perhaps a surprisingly good example here is Excel, which also, to some extent, keeps track of the chain of cause to effect in that you can view the source of the formula and see what other cells were used to compute the result. You do have to backtrace the causal chain manually by looking up the cells one by one. But it is the best example I can think of.}

@@

\says[\J]{True, so let's stop to think about notations for a bit. We can understand them better by charting the scope of change that can be achieved through a particular notation against the difficulty of using the notation. Figure~\ref{fig:sub1} shows this for a hypothetical accounting system and a spreadsheet system.}

@@

\says[\J]{Right. This is already better than regular software, because there are appropriate notations for changes at multiple difficulty levels, but it is not perfect...}

@@

\says[\J]{I do not have a perfect answer to this question, but you can think of the basic substrate as a sort of ``user interface physics'' within which specific visual notations for particular problems could be defined.\footnote{Constraint-based systems for specifying user interfaces, such as Cassowary \cite{borning-1997-constraints}, can be seen as related to this idea, although they focus more on rendering the layout for the end-user than on developing a foundational underlying substrate. Additionally, individual constraints are too low-level for a human-friendly UI substrate, although they could well be what such a substrate ``compiles'' to in order to run.} If we had that, then I think we would make a step towards the ideal substrate for programming. And the substrate needs to live within an actual running materialized system, rather than a static pre-scription of it like a program in source code today.}

@@

\says[\J]{Does that shed any light on why programming language researchers are so focused on formalization and proofs?}

@@

\says[\J]{I do not understand how you are supposed to make general and unifying claims if you do not first start by looking at the concrete and specific!}

\says[\T]{Indeed, and there are two examples of this that come to mind. For most of its history, particle physics consisted of two sub-cultures that Peter Galison refers to as the ``image'' and ``logic'' traditions \cite{galison-1997-image}. While the latter focused on statistical analysis of observations, the former attempted to capture ``golden events'' that show the decay of a single particle. These concrete traces went on to be hugely influential in shaping the theory of particle physics. And in the case of city planning, Jane Jacobs made a similar argument \cite{jacobs-1961-death} in favor of looking at specific cases, which she calls ``unaverage clues''. An example is a chain of bookshops that always remain open until late, except for one branch in Brooklyn. The clue tells you something significant about that part of the city.}

\says[\J]{OK, but on reflection, we need something in between. In order to do any formalisation or a user study, you need conceptual clarity on what you're studying and what you're trying to find out. I don't think we have any such clarity on many interesting aspects of programming systems.}

@@

\says[\T]{Interesting, this kind of reasoning seems to be quite common in software engineering. I can think of another example from a book about the history of anti-ballistic missile systems \cite{slayton-2013-arguments}. One of the arguments for why such systems cannot be reliably built was based on comparing the rate of change in the environment and the rate of change of the system. The idea was that the US cannot build a reliable defence system, because changing the software takes more work than changing some of the characteristics of the Soviet missiles. Ergo, their defences won't be able to keep pace with Soviet innovations in offensive capability. Again, this is a somewhat mathematical argument that provides conceptual clarity.}

\says[\J]{Right. I think we can refer to such models that provide conceptual clarity \emph{proto-theories}. I would say that we need more thinking like this! They can be later refined to more formal models, but not prematurely. I would also add that mathematical formulations like those above are not the only kinds of proto-theories we should be creating.}

@@

\says[\J]{The list is quite long and I do not think it is complete, but it includes some of the
  characteristics of programming systems that we have talked about already. For example,
  \emph{modes of interaction} captures whether you do everything in a single environment or
  whether ``using'' is distinct from ``programming''; \emph{notational structure} captures whether
  there is just a single adaptable notation or a range of different ones; \emph{self-sustainability}
  captures the degree to which the system can be modified from within itself and so on.}

@@

\says[\T]{Didn't you link the change from thinking about programming languages to thinking about programming systems to the idea of Kuhnian paradigm shift before?}

\says[\J]{Yes, it would be a corrective to the shift that Gabriel observed in the opposite direction\ \cite{rpg-2012-revolution}; not by \emph{fighting} PL research, but by simply making space to do the more general thing. To truly establish a new paradigm, then the shift to thinking about systems would put more emphasis on research methods that can talk about them in their full complexity. It would not \emph{invalidate} programming language theory, but it could make it more of a speciality instead of the default. However, we are quite far from what Kuhn called ``normal science''. It is still early days for programming systems research!}

@@

\says[\J]{Right. The full realisation of the aims of the framework is a long-term research program. What we have already is the seed to start it off. We have a set of dimensions that come with qualitative descriptions. Future work, as set out towards the end of my dissertation, involves making these descriptions more precise and possibly even quantifiable.}

@@

\says[\J]{That is certainly one task for the experimentalists. But they can also explore the design space to find new designs. To get back to our original motivation of making programming easier, my thesis explores programming systems that have three properties: \emph{notational freedom}, which would give you the kind of adaptable substrate we talked about,
\emph{explicit structure}, which is a pre-condition for building such a substrate, and
\emph{self-sustainability}, which would make it possible to evolve the system from
within itself, enabling a virtuous cycle of self-improvement.}

@@

\says[\T]{Well, Joel was practicing the elevator pitch for his PhD thesis, but we got a bit distracted. We ended up talking about whether it is possible for programming to be easier, whether there are any fundamental limits to that, and also what kind of research methodology can get us there!}

@@

\says[\J]{On a more general level, we agreed that we need more thinking about interactive, stateful programming systems and their characteristics. Programming can only become easier if we start thinking about how programming systems are used. In other words, think more about \emph{programming systems} than about \emph{programming languages}. Today, this is tricky, because we lack the right vocabulary for talking about programming systems, but the technical dimensions are a useful \emph{proto-theory} that serves as the starting point!}

@@

\says[\J]{Well, we do not have a concrete system yet, but we believe it will have a number of general properties. It needs to be built around a programming substrate that can grow with the user. It should make it easy to make small changes to the system, but also allow making more complex and difficult changes. We also need to think about programming as gradual adaptation, to overcome Brooks' ``doom prophecy'' of essential complexity. But I remain optimistic.}

@@

\says[\J]{I would love to see more research that looks at how programming languages are embedded in broader stateful interactive systems. This is something where we are lacking both theory and new interesting design ideas. I also think programming researchers should look at things that we do not really see as programming languages. What is it that makes systems like Excel, but also Flash, Hypercard or game engines work well as programming tools? We also need to be more fearless in looking at problems that are interesting, but informal. If we are to study informal properties that are not simply true or false, we will also need more focus on finding consensus among researchers working on programming tools, such as agreeing on the right proto-theories on which to further advance our work.\footnote{A recent seminar also focused on the problem of developing new ``theories of programming'' \cite{latoza-2023-theories}. However, we believe that a theory needs to start as an informal proto-theory and that \emph{premature} scientific framing in terms of hypotheses may inhibit progress.}}
